---
title: "Description and usage of MsBackendMgf"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendMgf}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMgf")`<br />
**Authors**: `r packageDescription("MsBackendMgf")[["Author"]] `<br />
**Last modified:** `r file.info("MsBackendMgf.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `Spectra` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports
interchangeable use of different *backends* to import MS data from a
variety of sources (such as mzML files). The `MsBackendMgf` package
allows the import of MS/MS data from mgf ([Mascot Generic
Format](http://www.matrixscience.com/help/data_file_help.html))
files. This vignette illustrates the usage of the `MsBackendMgf`
package.

# Importing MS/MS data from mgf files

Mgf files store one to multiple spectra, typically centroided and of
MS level 2. In our short example below, we load 2 mgf files which are
provided with this package. Below we first load all required packages
and define the paths to the mgf files.

```{r load-libs}
library(Spectra)
library(MsBackendMgf)

fls <- dir(system.file("extdata", package = "MsBackendMgf"),
           full.names = TRUE, pattern = "mgf$")
fls
```

MS data can be accessed and analyzed through `Spectra` objects. Below
we create a `Spectra` with the data from these mgf files. To this end
we provide the file names and specify to use a `MsBackendMgf()`
backend as *source* to enable data import.

```{r import}
sps <- Spectra(fls, source = MsBackendMgf())
```

With that we have now full access to all imported spectra variables
that we list below.


```{r spectravars}
spectraVariables(sps)
```

Besides default spectra variables, such as `msLevel`, `rtime`,
`precursorMz`, we also have additional spectra variables such as the
`title` of each spectrum in the mgf file.

```{r instrument}
sps$rtime
sps$title
```

In addition we can also access the m/z and intensity values of each
spectrum.

```{r mz}
mz(sps)
intensity(sps)
```

When importing a large number of mgf files, setting `nonStop = TRUE`
prevents the call to stop whenever problematic mgf files are
encountered. 

```{r all-import, eval = FALSE}
sps <- Spectra(fls, source = MsBackendMgf(), nonStop = TRUE)
```

# Session information

```{r}
sessionInfo()
```
