---
title: "Description and usage of MsBackendMgf"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendMgf}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{Spectra}
    %\VignetteDepends{Spectra,BiocStyle}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendMgf")`<br />
**Authors**: `r packageDescription("MsBackendMgf")[["Author"]] `<br />
**Last modified:** `r file.info("MsBackendMgf.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `Spectra` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports
interchangeable use of different *backends* to import MS data from a
variety of sources (such as mzML files). The `MsBackendMgf` package
allows the import of MS/MS data from mgf ([Mascot Generic
Format](http://www.matrixscience.com/help/data_file_help.html))
files. This vignette illustrates the usage of the `MsBackendMgf`
package.

# Importing MS/MS data from mgf files

Mgf files store one to multiple spectra, typically centroided and of
MS level 2. In our short example below, we load 2 mgf files which are
provided with this package. Below we first load all required packages
and define the paths to the mgf files.

```{r load-libs}
library(Spectra)
library(MsBackendMgf)

fls <- dir(system.file("extdata", package = "MsBackendMgf"),
           full.names = TRUE, pattern = "mgf$")
fls
```

MS data can be accessed and analyzed through `Spectra` objects. Below
we create a `Spectra` with the data from these mgf files. To this end
we provide the file names and specify to use a `MsBackendMgf()`
backend as *source* to enable data import.

```{r import}
sps <- Spectra(fls, source = MsBackendMgf())
```

With that we have now full access to all imported spectra variables
that we list below.


```{r spectravars}
spectraVariables(sps)
```

Besides default spectra variables, such as `msLevel`, `rtime`,
`precursorMz`, we also have additional spectra variables such as the
`TITLE` of each spectrum in the mgf file.

```{r instrument}
sps$rtime
sps$TITLE
```

By default, fields in the mgf file are mapped to spectra variable names using
the mapping returned by the `spectraVariableMapping()` function:

```{r spectravariables}
spectraVariableMapping()
```

The names of this `character` vector are the spectra variable names (such as
`"rtime"`) and the field in the mgf file that contains that information are the
values (such as `"RTINSECONDS"`). Note that it is also possible to overwrite
this mapping (e.g. for certain mgf *dialects*) or to add additional
mappings. Below we add the mapping of the mgf field `"TITLE"` to a spectra
variable called `"spectrumName"`.

```{r map}
map <- c(spectrumName = "TITLE", spectraVariableMapping())
map
```

We can then pass this mapping to the `backendInitialize` method, or the
`Spectra` constructor.

```{r import2}
sps <- Spectra(fls, source = MsBackendMgf(), mapping = map)
```

We can now access the spectrum's title with the newly created spectra variable
`"spectrumName"`:

```{r spectrumName}
sps$spectrumName
```

In addition we can also access the m/z and intensity values of each
spectrum.

```{r mz}
mz(sps)
intensity(sps)
```

The `MsBackendMgf` backend allows also to export data in mgf format. Below we
export the data to a temporary file. Note that we use again our custom mapping
of variables such that the spectra variable `"spectrumName"` will be exported as
the spectrums' title.

```{r export}
fl <- tempfile()
export(sps, file = fl, mapping = map)
```

We next read the first lines from the exported file to verify that the title was
exported properly.

```{r export-check}
readLines(fl)[1:12]
```

Note that we can also use the `MsBackendMgf` to export data from a `Spectra`
that uses a different backend to a file in mgf format. For this we change first
the backend of our `Spectra` object to `MsBackendDataFrame`, a backend that
keeps all data in memory. Subsequently we add also an additional spectra
variable we might want to export to the mgf file too.

```{r}
sps2 <- setBackend(sps, MsBackendDataFrame())
sps2$new_variable <- "A"
```

Directly calling `export` on `sps2` would throw an error because the
`MsBackendDataFrame` does not provide an `export` method. Thus we tell the
function to use the export method from the `MsBackendMgf` by setting `backend =
MsBackendMgf()`.

```{r}
export(sps2, backend = MsBackendMgf(), file = fl)
readLines(fl)[1:12]
```

We can see that also our newly defined variable was exported. Also, because we
did not provide our custom variable mapping this time, the variable
`"spectrumName"` was **not** used as the spectrum's title.

# Session information

```{r}
sessionInfo()
```
